ML Estimation of Matrix--$t$ Distributions with Expectation Maximization

### 2017-01-12 paper.tex
Finished editing section 2.1.

We first review some basic definitions on matrix--$t$ distributions. A matrix--$t$ distribution $T_{p,m}(M,\Sigma,\Omega,\nu)$ of dimensions $p\times m$ is specified by 4 parameters: the location matrix $M$ of dimensions $p\times m$ , the row and column SPD scale matrices $\Sigma$ and $\Omega$ of dimensions $p\times p$ and $m\times m$, respectively, and the degrees of freedom parameter $\nu>0$. A $p \times m$ random matrix $T$ is said to be matrix--$t$ distributed $T_{p,m}(M,\Sigma,\Omega,\nu)$ with location parameter $M$ if given the scale parameter $S$, a SPD matrix of dimensions $p\times p$, the random matrix $T$ is matrix normal distributed $N_{p,m}(M,S,\Omega)$ and $S$ is itself inverse Wishart distributed $W^{-1}(\Sigma,\nu +p -1)$:
\begin{equation}
T|M,\Sigma,\Omega,\nu,S \sim N_{p,m}(M,S,\Omega)\label{eq:tdef1}
\end{equation}
and
\begin{equation}
S|\Sigma,\nu \sim W^{-1}(\Sigma,\nu+p-1).\label{eq:tdef2}
\end{equation}

The matrix normal distribution $N_{p,m}(M,S,\Omega)$ has density
\begin{equation}
\frac{1}{(2\pi)^{pm/2}|\Omega|^{p/2}|S|^{m/2}}\exp\left(-\frac{1}{2}\textrm{tr}\left[\Omega^{-1}(X-M)^{\textrm{t}}S^{-1}(X-M)\right]\right)\label{eq:ndistpdf}
\end{equation}
at $X$ while the inverse Wishart distribution $W^{-1}(\Sigma,n)$ with $n>p-1$ has density
\begin{equation}
\frac{|\Sigma|^{\frac{n}{2}}}{2^{\frac{n p}{2}}|S|^{\frac{n+p+1}{2}}\Gamma_{p}(\frac{n}{2})}\exp\left(-\frac{1}{2}\textrm{tr}\left[\Sigma S^{-1}\right]\right)\label{eq:wdistpdf}
\end{equation}
at $S$. Here $\Gamma_{p}(\cdot)$ is the
multivariate gamma function defined in Eq.~\eqref{eq:multgammadef}.

Follow the definition in Eqs.~\eqref{eq:tdef1} and \eqref{eq:tdef2} and the densities given above we can see the density function $f(T;\left\{M,\Sigma,\Omega,\nu\right\})$ of a matrix--$t$ distribution $T_{p,m}(M,\Sigma,\Omega,\nu)$ satisfies
\begin{equation}
\frac{\Gamma_{p}\left(\frac{p+\nu+m-1}{2}\right)}{\Gamma_{p}\left(\frac{\nu+p-1}{2}\right)\pi^{\frac{mp}{2}}|\Omega|^{\frac{p}{2}}|\Sigma|^{\frac{m}{2}}}
|I_p+\Sigma^{-1}(T-M)\Omega^{-1}(T-M)^{\textrm{t}}|^{\frac{1-m-p-\nu}{2}}
\label{eq:matrixtpdf}
\end{equation}
with $I_p$ the identity matrix of dimension $p$. The marginalization over $S$ can be obtained by using the Jacobain determinant
\begin{equation}
...
\end{equation}
and the integral identity \citep{gupta1999matrix}
\begin{equation}
\int_{\left\{S_p\succ 0\right\}}e^{-\textrm{tr}[S_pA]}|S_p|^{\alpha}\,dS_p=\Gamma_{p}\left(\alpha+\frac{1+p}{2}\right)|A|^{\alpha+\frac{1+p}{2}},\label{eq:intmultgammadef}
\end{equation}
where $A$ is a $p\times p$ nonsingular matrix.

We repeat the results here because, along the way, we will utilize several results that are useful in the derivation of the EM formulae.

We will show that the matrix
\begin{equation}
T=(S^{-\frac{1}{2}})^{t}X+M\label{eq:change}
\end{equation}
follows a matrix--$t$ distribution with degrees of freedom $n=\nu-p+1$. Here $S^{\frac{1}{2}}$ is the lower Cholesky factor of $S$, i.e., $S=S^{\frac{1}{2}}(S^{\frac{1}{2}})^t$. The Jacobian determinant of the transformation \eqref{eq:change} is
\begin{equation}
\left|\frac{\partial T}{\partial X}\right|=|S|^{-\frac{m}{2}}.\label{eq:jac}
\end{equation}
Using Eq.~\eqref{eq:ndistpdf}, Eq.~\eqref{eq:wdistpdf} and Eq.~\eqref{eq:jac} we can see the joint density of $T$ and $S$ is given as
\begin{equation}
\begin{split}j(T,S;\left\{M,\Sigma,\Omega,n\right\})=\frac{\pi^{-\frac{mp}{2}}|\Sigma|^{\frac{n+p-1}{2}}|\Omega|^{\frac{n+m-2}{2}}|S|^{\frac{m+n-2}{2}}}{2^{p\frac{n+m+p-1}{2}}\Gamma_{p}(\frac{n+p-1}{2})}\times\\
\exp\left(-\frac{1}{2}\textrm{tr}\left[S(\Sigma+(T-M)\Omega^{-1}(T-M)^{\textrm{t}}\right]\right).
\end{split}
\label{eq:jointpdf}
\end{equation}

Finally we obtain the density \eqref{eq:matrixtpdf} of the matrix--$t$ distribution
by integrating over $S$ in the above expression, using Eq.~\eqref{eq:change} and slightly
rearranging the terms. In particular noting that
$|\Sigma+(T-M)\Omega^{-1}(T-M)^{\textrm{t}}|=|\Sigma||I_p+\Sigma^{-1}(T-M)\Omega^{-1}(T-M)^{\textrm{t}}|$.

>>>>

\subsection{Matrix--$t$ distribution}

A matrix--$t$ distribution $T_{p,m}(M,\Sigma,\Omega,\nu)$ of dimensions $p\times m$ is specified by 4 parameters: the location matrix $M$ of dimensions $p\times m$ , the row and column SPD scale matrices $\Sigma$ and $\Omega$ of dimensions $p\times p$ and $m\times m$, respectively, and the degrees of freedom parameter $\nu>0$. A $p \times m$ random matrix $T$ is said to be matrix--$t$ distributed if given the scale parameter $S$, a SPD matrix of dimensions $p\times p$, the matrix $T$ is matrix normal distributed $N_{p,m}(M,S,\Omega)$ and $S$ is itself inverse Wishart distributed $W^{-1}(\Sigma,\nu +p -1)$:
\begin{equation}
T|M,\Sigma,\Omega,\nu,S \sim N_{p,m}(M,S,\Omega)\label{eq:tdef1}
\end{equation}
and
\begin{equation}
S|\Sigma,\nu \sim W^{-1}(\Sigma,\nu+p-1).\label{eq:tdef2}
\end{equation}
The matrix normal distribution $N_{p,m}(M,S,\Omega)$ has density
\begin{equation}
\frac{1}{(2\pi)^{pm/2}|\Omega|^{p/2}|S|^{m/2}}\exp\left(-\frac{1}{2}\textrm{tr}\left[\Omega^{-1}(X-M)^{\textrm{t}}S^{-1}(X-M)\right]\right)\label{eq:ndistpdf}
\end{equation}
at $X$ while the inverse Wishart distribution $W^{-1}(\Sigma,n)$ with $n>p-1$ has density
\begin{equation}
\frac{|\Sigma|^{\frac{n}{2}}}{2^{\frac{n p}{2}}|S|^{\frac{n+p+1}{2}}\Gamma_{p}(\frac{n}{2})}\exp\left(-\frac{1}{2}\textrm{tr}\left[\Sigma S^{-1}\right]\right)\label{eq:wdistpdf}
\end{equation}
at $S$. Here $\Gamma_{p}(\cdot)$ is the
multivariate gamma function defined in Eq.~\eqref{eq:multgammadef}.

Follow the definition in Eqs.~\eqref{eq:tdef1} and \eqref{eq:tdef2} and the densities given above we can see the density function $f(T;\left\{M,\Sigma,\Omega,\nu\right\})$ of a matrix--$t$ distribution $T_{p,m}(M,\Sigma,\Omega,\nu)$ satisfies
\begin{equation}
\frac{\Gamma_{p}\left(\frac{p+\nu+m-1}{2}\right)}{\Gamma_{p}\left(\frac{\nu+p-1}{2}\right)\pi^{\frac{mp}{2}}|\Omega|^{\frac{p}{2}}|\Sigma|^{\frac{m}{2}}}
|I_p+\Sigma^{-1}(T-M)\Omega^{-1}(T-M)^{\textrm{t}}|^{\frac{1-m-p-\nu}{2}}
\label{eq:matrixtpdf}
\end{equation}
with $I_p$ the identity matrix of dimension $p$. The marginalization over $S$ can be obtained by using the Jacobain determinant for inverse SPD matrix $U=S^{-1}$
\begin{equation}
\left|\frac{\partial U}{\partial S}\right|=|S|^{-p-1}
\end{equation}
and the integral identity \citep{gupta1999matrix}
\begin{equation}
\int_{\left\{S_p\succ 0\right\}}e^{-\textrm{tr}[S_pA]}|S_p|^{\alpha}\,dS_p=\Gamma_{p}\left(\alpha+\frac{1+p}{2}\right)|A|^{-\alpha-\frac{1+p}{2}},\label{eq:intmultgammadef}
\end{equation}
where $A$ is a $p\times p$ nonsingular matrix.

Similar as matrix normal distributions, matrix--$t$ distributions are defined up to a multiplicative constant, i.e., $T_{p,m}(M,c\Omega,\Sigma/c,\nu)$ has the same distribution as $T_{p,m}(M,\Omega,\Sigma,\nu)$ for any $c>0$.
